---
title: "How emerging economies are leapfrogging fossil fuels"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
---

# Introduction

Today, the cheapest way to power a society, is green, clean, renewable electricity. We'll show the "fossil fuel detour" taken by the West, and how some of the poorest countries in the world are building better from the start.  

*In this repository, you will find the methodology, data and code behind the stories that came out of this analysis.*

**Watch the video here:** [Youtube | DW Planet A |  The unexpected way India is beating the U.S.](https://www.youtube.com/watch?v=qoNr6hjy7aE)

**Story by:** [Kira Schacht](https://www.dw.com/en/kira-schacht/person-46893544)

- `analysis.Rmd` is the main file that contains the R code for the analysis
- `data/...` contains the raw data files used in the analysis

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/SchachtK/Library/CloudStorage/OneDrive-DeutscheWelle/_Projects/2604 leapfrogging/extern') #INSERT YOUR PROJECT DIRECTORY HERE

options(scipen=999)

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

needs(jsonlite, tidyverse, readxl, ggbeeswarm,ggtern)

```


# Read data

## UN Country metadata

- [UN Country isocodes and continent/regional affiliations](https://unstats.un.org/unsd/methodology/m49/overview/) 
- [World Bank income groups](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups  )

```{r}
meta_region = read.csv("data/meta/UNSD — Methodology.csv", sep = ";") %>% 
  select(countrycode = ISO.alpha3.Code,
         country = Country.or.Area,
         region = Region.Name, subregion = Sub.region.Name,
         LDC = Least.Developed.Countries..LDC.) %>% 
  mutate(LDC = (LDC == "x"))
meta_income = read_xlsx("data/meta/CLASS_2026_07_01.xlsx") %>% 
  select(countrycode = Code, income = `Income group`)

meta = full_join(meta_region, meta_income, by = "countrycode")

rm(meta_region, meta_income)
```

## GDP data

[Penn World Table version 11.0](https://www.rug.nl/ggdc/productivity/pwt/?lang=en): Output-side real GDP at chained PPPs (in mil. 2021US$) per capita


```{r}
gdp = read_xlsx('data/pwt110.xlsx', sheet = "Data") %>% 
  select(countrycode, country, year, rgdpo, pop) %>% 
  group_by(countrycode) %>%
  mutate(gdp_pc = rgdpo / pop,
         g10m = last(pop, order_by = year) >= 10) #countries with at least 1m people in 2023
#pop in mil
```


## Historical energy data

[Primary, Final and Useful Energy Database (PFUDB)](https://iiasa.ac.at/models-tools-data/pfudb): Final energy consumption by source (TJ/y) and population (1000 persons), for select regions & countries.

```{r}
l = list.files("data/pfu")
#merge excel files in this folder
pfu = map(l, ~ read_xlsx(paste0("data/pfu/", .x), sheet = "data")) %>%
  list_rbind() %>% 
  select(-Notes, -Scenario, -Model, -Unit) %>% 
  filter(!(is.na(Variable))) %>% 
  pivot_longer("1900.0":"2014.0", names_to = "year", values_to = "value") %>% 
  mutate(Variable = str_replace_all(Variable, "FEn.Fuel.|Other variables.|.Sector.All Sectors.All Uses", "") %>% 
           str_replace("All Fuels", "Total"),
         year = as.integer(year)) %>% 
  filter(!(Variable %in% c("GDP","Population")))
rm(l)
```


## IEA Energy Data

- Energy consumption by fuel
- Electricity generation by fuel

```{r read-data}

#electricity generation data
d_elec = fromJSON('data/IEA/ElecGenByFuel.json') %>%
  rename(flow = product, flowLabel = productLabel, product = flow, productLabel = flowLabel)

# Other stats
d = fromJSON('https://api.iea.org/stats?series=BALANCES') %>% 
  filter(flow %in% c("TFC","ELECTR","INDPROD","IMPORTS","TES")) %>% #keep only relevant flows
  full_join(d_elec) %>% #add electricity data
  mutate(year = as.integer(year),
         country = ifelse(country == "GUYANA", "GUY", country)) %>% # Fix Guyana country code
  select(-flowOrder, -seriesLabel) %>%
  rename(countrycode = country, country = short)# %>% 
  #filter(flow %in% flows, year < 2024) #2024 no gdp data yet
rm(d_elec)


```

## Merge PFU data with IEA data

Check how similar the two datasets are:

```{r}
#Compare reported energy consumption totals between IEA and PFU data for select countries
tmp = pfu %>%
  filter(Variable == "Total",
         Region %in% c("United States","China","Germany","India")) %>%
  rename(country = Region) %>%
  select(-2) %>% mutate(source = "PFU")

tmp_test = d %>%
  filter(flow == "TFC", product == "TOTAL",
         countrycode %in% c("USA","CHN","DEU","IND")) %>% 
  mutate(country = case_when(countrycode == "USA" ~ "United States",
                               countrycode == "CHN" ~ "China",
                               countrycode == "DEU" ~ "Germany",
                               countrycode == "IND" ~ "India")) %>%
  select(1,2,4) %>% 
  mutate(source = "IEA") %>% 
  bind_rows(tmp)
rm(tmp)

ggplot(tmp_test,
       aes(x = year, y = value, 
           linetype = source, color = country)) +
  geom_line() +
  labs(x = "Year", y = "Total final energy consumption (TJ/y)") + theme_minimal()
#look very similar, nice!
rm(tmp_test)
```


```{r}
#make equivalent variables
pfu_vars = tibble(
  pfu = pfu$Variable %>% unique %>% sort,
  productLabel = c("Biofuels and waste",
                   "Coal and coal products",
                   "Electricity",
                   "Solar, wind and other renewables",
                   "Heat",
                   "Hydropower",
                   "Natural gas",
                   "Nuclear",
                   "Other",
                   "Oil and oil products",
                   "Solar, wind and other renewables",
                   "Total",
                   "Solar, wind and other renewables")
)


pfu_merge = pfu %>%
  #append country information from gdp data (bc country names match better)
  left_join(gdp %>% group_by(countrycode,country) %>% summarize,
            by = c("Region" = "country")) %>% 
  mutate(countrycode = ifelse(is.na(countrycode) & Region == "OECD-90", "OECDTOT", countrycode)) %>%
  #merge variable names with IEA data
  left_join(pfu_vars, by = c("Variable" = "pfu")) %>% 
  select(-Variable) %>% 
  #group by fuel type so only one entry per
  #bc solar, wind and other renewables have multiple entries
  group_by(across(-value)) %>% 
  summarize(value = sum(value, na.rm=T)) %>%
  #append missing metadata from IEA dataset
  left_join(d %>% filter(flow == "TFC") %>% group_by(flow, flowLabel, units, product, productLabel, productOrder) %>% summarize,
            by = "productLabel") %>% 
  ungroup %>% select(-Region)

d = bind_rows(d %>% mutate(source = "IEA"),
              pfu_merge %>% mutate(source = "PFU"))

rm(pfu_merge, pfu_vars)
```

## Merge energy data with GDP data, add country metadata

```{r}
d = d %>% select(-country) %>% 
  left_join(gdp %>% select(-country),
            by = c("countrycode", "year")) %>% 
  left_join(meta, by = c("countrycode")) %>% 
  #use country names from metadata, add taiwan label
  mutate(country = ifelse(countrycode == "TWN", "Taiwan", country)) %>%
  relocate(country, .before = year)

```


## Assign electricity and energy types to classes

```{r}
renew = c("ELECTR","GEOTHERM","HYDRO",
          "ESOLARPV","ETIDE", "EHGEOTHERM", "EHSOLARTH", "EHYDRO","EWIND")
fossil = c("COAL","MTOTOIL","NATGAS","EHCOAL", "EHNATGAS", "EHOIL","CRNGFEED","TOTPRODS")
other = c("NUCLEAR","COMRENEW","HEAT","EHNUCLEAR", "EHOTHER","EHWASTE","EHBIOMASS")
#electr biofuels renewable? 

d_types = d %>% filter(flow %in% c("ELECTR","TFC")) %>%
  group_by(flow, product, productLabel) %>% 
  summarize() %>% 
  mutate(type = case_when(product %in% fossil ~ "fossil",
                          product %in% renew ~ "renewable",
                          product %in% other~ "other"))

rm(renew,fossil,other)

d_types %>% group_by(type) %>%
  summarize(productLabel = paste(productLabel, collapse =", ")) %>% 
  knitr::kable()
```


# Energy consumption

## How did countries' energy mixes change over time?

Total final consumption (TFC) by source over time

```{r}
tfc_source = d %>%
  #filter for TFC, exclude total and sub-categories, use IEA data from 1990
  filter(flow == "TFC",
         !(product %in% c("TOTAL", "TOTPRODS","CRNGFEED")),
         #!is.na(country),
         #post-1990 data only from IEA
         !(year >= 1990 & source == "PFU")) %>% 
  #attach fuel type categories
  left_join(d_types, by = join_by(flow, product, productLabel)) %>% 
  group_by(countrycode, year, source) %>%
  mutate(share = value/sum(value)) %>% 
  select(1:2,4,share, type, 7:8,10:11,13,16:19)

cols = c("COAL" = "#9e2424", "NATGAS" = "#f7ad6d",
         "MTOTOIL" = "#eb6565", "ELECTR" = "#50cffa",
         "COMRENEW" = "#0044b8", "GEOTHERM" = "#0083e4",
         "HYDRO" = "#05b2fc","NUCLEAR" ="#D8DDE2","HEAT" ="#9099A3")
```

### Baseline countries: USA, India, China, OECD

Energy consumption shares: Fossil vs biomass vs electricity&renewable
```{r}
tfc_plot_baseline = tfc_source %>%
  filter(countrycode %in% c("USA","CHN","IND","OECDTOT")) %>% 
  #move biomatter into "other" category
  mutate(country = ifelse(countrycode == "OECDTOT", "OECD-90", country)) %>%
  group_by(countrycode, country, year, type) %>%
  summarise(share = sum(share)) %>% 
  pivot_wider(names_from = type, values_from = share, values_fill = 0) %>%
  rename(bio_other = other,
         electr_renew = renewable)

```


```{r}
ggtern(tfc_plot_baseline,
       aes(x = electr_renew, y = fossil,z = bio_other,
           color = country, fill = country)) +
  geom_path(size = 2) +
  geom_point(data = tfc_plot_baseline %>%
               group_by(country) %>% slice_max(year),
             size = 3, shape = 21, color = "white") +
  labs(z = "Biofuels\n& others",
       y = "Fossil fuels",
       x = "Electricity\n& renewables",
       title = "Emerging economies are skipping the fossil fuel detour",
       subtitle = "% of final energy consumption by source") +
  theme(axis.title=element_text(size=8)) +
  theme_rotate(degrees = 180)

```


```{r}
tfc_plot_income = tfc_source %>%
  #remove 3 outliers in low-income countries, limit to years where most countries have data
  filter(year >= 1990, year <= 2023, !is.na(income),
         !(countrycode %in% c("PRK","SYR","YEM"))) %>% 
  group_by(income, year, type) %>%
  summarise(value = sum(value, na.rm = T)) %>% 
  group_by(income, year) %>% 
  mutate(share = value/sum(value, na.rm = T)) %>%
  select(-value) %>% 
  pivot_wider(names_from = type, values_from = share, values_fill = 0) %>%
  rename(bio_other = other,
         electr_renew = renewable)

```


```{r}
#scatter chart
ggtern(tfc_plot_income,
       aes(x = electr_renew, y = fossil, z = bio_other, color = income)) +
  #add individual baseline countries
  geom_path(data = tfc_plot_baseline, aes(color = country), alpha = .8) +
  #lines and points by income group
  geom_path(size = 1.5) +
  geom_point(data = tfc_plot_income %>%
               group_by(income) %>% slice_max(year)) +
  labs(x = "Electricity\n&renewables",
       y = "Fossil fuels", y = "Bio\n&other",
       title = "Emerging economies are skipping the fossil fuel detour",
       subtitle = "% of final energy consumption by source") +
  theme(axis.title=element_text(size=8)) +
  theme_rotate(degrees = 180)
```


# Electricity mix over time

```{r}
d_elec = d %>%
  filter(flow == "ELECTR") %>% 
  group_by(countrycode) %>% 
  left_join(d_types, by = join_by(flow, product, productLabel)) %>% 
  group_by(countrycode, year) %>%
  mutate(share = value/sum(value))

cols = c("EHCOAL" = "#9e2424", "EHNATGAS" = "#f7ad6d", "EHOIL" = "#eb6565",
         "ESOLARPV" = "#50cffa", "ETIDE" = "#0044b8", "EHGEOTHERM" = "#0083e4",
         "EHSOLARTH" = "#05b2fc", "EHYDRO" = "#002186","EWIND" = "#081336",
         "EHNUCLEAR" ="#D8DDE2", "EHOTHER" = "#BFC7CE","EHWASTE" ="#9099A3","EHBIOMASS" = "#6D7780")
colsLabel = c("Coal" = "#9e2424", "Natural gas" = "#f7ad6d", "Oil" = "#eb6565",
         "Solar PV" = "#50cffa", "Tide" = "#0044b8", "Geothermal" = "#0083e4",
         "Solar thermal" = "#05b2fc", "Hydropower" = "#002186","Wind" = "#081336",
         "Nuclear" ="#D8DDE2", "Other" = "#BFC7CE","Waste" ="#9099A3","Biofuels" = "#6D7780")
```


## Current electricity mix by country


```{r}
elec_shares_current = d_elec %>% ungroup %>% 
  filter(!(countrycode %in% c("PRK","SYR","YEM")), year < 2024) %>% 
  group_by(country) %>% 
  slice_max(year) %>% 
  group_by(country, income, type) %>% 
  summarize(share = sum(share)) %>%
  pivot_wider(names_from = type, values_from = share, values_fill = 0)

elec_shares_current %>%
  group_by(income) %>%
  summarize(renewable = median(renewable))
```


## Column chart of current electricity sources:

Income group averages, China, India, Vietnam, US, Global average
```{r}
#global average
d_tmp_global = d_elec %>% ungroup %>% 
  filter(countrycode == "WORLD") %>%
  slice_max(year) %>% 
  mutate(group = "Global average") %>% select(group, productLabel, type, value, share)
#by income group
d_tmp_income = d_elec %>%
  filter(!is.na(income), year < 2024) %>% 
  group_by(group = income, year, type, productLabel) %>% 
  summarise(value = sum(value, na.rm = T), .groups = "drop") %>% 
  slice_max(year) %>% 
  group_by(group) %>% select(-year) %>% 
  mutate(share = value/sum(value))

#merge with individual countries
d_plot = d_elec %>% ungroup %>%
  filter(countrycode %in% c("USA","CHN","VNM","IND"), year < 2024) %>% 
  group_by(country) %>% 
  slice_max(year) %>% 
  select(group = country, productLabel, type, value, share) %>%
  bind_rows(d_tmp_global, d_tmp_income) %>% 
  #group Geothermal,Other sources,Solar thermal,Tide,Waste into other, since their shares are so low
  mutate(productLabel = ifelse(productLabel %in% c("Geothermal","Other sources","Solar thermal","Tide","Waste"), "Other",productLabel),
         type = ifelse(productLabel == "Other","other",type)) %>% 
  group_by(group, type, productLabel) %>%
  summarize(share = sum(share)) %>% ungroup %>% 
  #change order of countries for the chart
  mutate(group = fct_reorder2(group,
                              .x = share, .y = type,
                               .fun = function(x, y){-sum(x[y == "renewable"])}))

rm(d_tmp_global, d_tmp_income)
```


```{r}
ggplot(d_plot,
       aes(group, share, group = type, fill = productLabel)) +
  geom_col() +
  scale_fill_manual(values = colsLabel) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Year", y = "Share of electricity generation") +
  guides(fill = guide_legend(position = "top", direction = "horizontal")) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

```

# Quality control disclaimer

The code used in this analysis was written by the author. For all of our
analyses, if any portion of the code is LLM-generated, this is flagged
in the script.

All code in this project has been reviewed pre-publication by both the
author as well as a second person to ensure quality.

Should you have any questions or notice errors or inconsistencies,
please reach out to
[data-team\@dw.com](mailto:data-team@dw.com){.email}.

